### 2023-04-13 submission/manuscript.tex
Editing through Introduction

\begin{verbatim}
## Warning: package 'ggplot2' was built under R version 4.2.3
\end{verbatim}

\begin{verbatim}
## Warning: package 'tibble' was built under R version 4.2.3
\end{verbatim}

\begin{verbatim}
## Warning: package 'dplyr' was built under R version 4.2.3
\end{verbatim}

\hypertarget{waste-not-want-not-revisiting-the-analysis-that-called-rarefaction-into-question}{%
\section{Waste not, want not: Revisiting the analysis that called
rarefaction into
question}\label{waste-not-want-not-revisiting-the-analysis-that-called-rarefaction-into-question}}

@@

Since the development of sequencing technologies such as those provided
by 454 and Illumina, microbiome researchers have struggled to produce a
consistent number of sequences from each sample in a dataset. It is
common to observe more than 10-fold variation in the number of sequences
per sample {[}XXXXX{]}. Regardless of the source of this variation,
researchers desire approaches to control for uneven sampling effort. Of
course, this desire is not unique to microbiome research and is a
challenge faced by all community ecologists. Common approaches to
controlling uneven sampling efforts have included use of proportional
abundance (i.e., relative abundance), normalization of counts, parameter
estimation, and rarefaction.

In 2014 Paul McMurdie and Susan Holmes published their ``Wast not, want
not: why rarefying microbiome data is inadmissible'' (WNWN) in PLOS
Computational Biology {[}XXXXX{]}. This paper has had a significant
impact on the approaches that microbiome researchers use to analyze 16S
rRNA gene sequence data. According to Google Scholar, this paper has
been cited more than 2,300 times as of January 2023. Anecdotely, I have
received correspondence from researchers over the past 10 years asking
how to address critiques from reviewers who criticize my correspondents'
analysis for rarefying (e.g., see
\href{https://twitter.com/inanna_nalytica/status/1264679859672006656}{this
Twitter thread}). I have also received these types of comments from
reviewers, specifically in regards to a preprint that I posted in 20XX
in regards to my critique of the practice of removing rare taxa from
analyses {[}XXXXX{]}. In the process of responding to these critiques
and preparing a manuscript investigating rarefaction and other
approaches to control for uneven sequencing effort, I decided to
reassess the WNWN study including their definitions, simulations, and
analyses.

@@

As I attempted to reproduce the results of WNWN, I noticed that the step
that purported to rarefy the data only performed one subsampling of the
data (Lines 404 through 416 of
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}).
This caused me to re-inspect how McMurdie and Holmes defined
``rarefying'' in the following quoted text from their paper:

@@

It was unfortunate that McMurdie and Holmes used the term ``rarefying''
here and throughout their manuscript. The authors were correct to state
that the distinction between ``rarefying'' and ``rarefaction'' is
confusing and leads to their conflation. In my experience, subsequent
researchers have conflated the results of this study of the effects of
rarefying data with rarefaction of data. As an example, Willis (XXXXX)
describes problems with rarefaction rather than rarefying data when
citing WNWN in her paper proposing alternatives to rarefaction for use
with alpha diversity data:

@@

Adding to the confusion is that the papers cited in the first sentence
of the quote I WNWN included above either do not use the words
``rarefy'' or ``rarefying'' or use them interchangably with
``rarefaction''. In hindsight, as shown in the quoted text, McMurdie and
Holmes do emphasize the distinction between rarefying and rarefaction.
However, because they seem to have coined a new meaning for rarefying,
they seem to have only added to the confusion by using the generally
used verb form of rarefaction. Further confusion comes from the author's
admonition in the final sentence that some researchers have failed to
repeat the subsampling step. To most scientists, repeating the
subsampling step is rarefaction. My preference is to use subsampling as
the term describing the process they refer to as rarefying. In other
words rarefaction with a single randomization.

To provide a more clear definition of rarefaction, I propose the
following:

@@

This definition aligns well with how rarefaction was originally defined
for comparing richness (i.e., the number of taxa in a community) across
communities when communities are sampled to different depths. It is
important to note that this procedure generates substantially different
results to those obtained without accounting for uneven sampling effort
or using relative abundances, normalized counts, compositional data
transformatins, variance stabilization procedures, or estimation
techniques. I have explored the differences in results obtained using
the diversity of approaches for controlling for uneven sampling effort;
rarefaction, as described here, outperforms the other approaches
{[}XXXXX{]}.

With this more general approach to rarefaction, rarefaction can be
performed using any alpha or beta diversity metric. This strategy has
been widely used by my research group and others. The procedure outlined
above could also be used for hypothesis tests of differential abundance;
however, thought would need to be given to how to synthesize the results
of these tests across a large number of replications.

@@

McMurdie and Holmes analyzed the effect of rarefying and other
approaches on clustering accuracy using what they called ``Simulation
A'' in their Figure 2A and elsewhere in their paper. In Simulation A,
they investigated the ability to correctly assign samples to one of two
clusters using simulated data used to generate 40 samples from each of
two distributions. A variety of approaches were used to calculate
distances between the samples those distances were used as input to
partitioning around mediods (PAM) clustering with two groups. The
accuracy of the cluster assignment was used as the metric to assess
performance. This analysis was performed in the
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
R-flavored markdown file that was published as Protocol S1 in the
original paper. Line numbers from this file will be refernced with an
``L'' as a prefix.

The two distributions were generated using human fecal and ocean data
originally take from the GlobalPaterns dataset (L129) {[}XXXXX{]}. To
generate a fecal and ocean template distribution, the authors included
any operational taxonomic unit (OTU) that appeared in more than one of
the 4 fecal and 3 ocean samples (L60 and L137). The OTUs were sorted by
how many of the 7 samples the OTUs were observed in followed by their
total abundance across all 7 samples (L139). From this sorted list they
identified the identifiers of the first 2000 OTUs (L66). Returning to
the 7 samples they selected the 2000 most common and abundant OTUs and
pooled the abundances of the fecal and ocean samples separately to
create two templates (L144, L159-160, L197-198).

Next, the fecal and ocean templates were mixed in 8 different fractions
to generate two community types that differend by varying effect sizes
(1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220). To simulate
the variation in sequencing depth across the 80 samples, they normalized
he number of sequences from each of the \texttt{gp\_n\_samples} samples
in the GlobalPatterns dataset so that the median number of sequences
(N\textsubscript{L}) for the GlobalPatterns had 1,000, 2,000, 5,000, or
10,000 sequences (L324-325). They then randomly sampled the
\texttt{gp\_n\_samples} normalized sequencing depths to generate 80
sampling depths. From each community type, they simulated 40 samples by
sampling to the desired number of reads (L73, L230-233 and L326-327).
Each simulation condition was repeated 5 times (L85). This resulted in
160 simulations (8 effect sizes x 4 median sampling depths x 5
replicates = 160 simulations). Finally, they removed rare and low
prevalence OTUs in two steps. First, they removed any OTUs whose total
abundance was less than 3 across all 80 samples and that did not appear
in at least 3 samples (L368-386). Second, they removed any OTUs that did
not have more than 1 sequence in more than 5\% of the 80 samples (i.e.,
4 samples) and that did not have a total abundance across the 80 samples
greater than one half of the number of samples in each community type
(i.e., 20) (L523-538, L551).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Simulated conditions were only replicated 5 times each, which caused
  results to be sensitive to random number generator seed
\item
  The average sizes of the libraries were small by modern standards
\item
  DeSeq-based variance stabilization was used with inappropriate
  distance calculation methods
\item
  A single subsampling of each dataset was evaluated rather than using
  rarefaction, which likely resulted in noisier data
\item
  Results using PAM clustering were not directly compared to those of
  K-means and hierarchical clustering
\item
  Subsampling removed the smallest 15\% of the samples, which penalized
  accuracy values by 15 percentage points
\item
  The distribution of library sizes was poorly chosen
\item
  A filtering step was applied to remove rare taxa from the simulated
  datasets, which could distort the shape of the communities
\item
  No test of effect of transformation methods to account for effects of
  uneven sample size between treatment groups
\item
  Clustering accuracy was used rather than direct comparisons beta
  diversity
\item
  No consideration of effects of transformations on alpha diversity
  metrics
\end{enumerate}

@@

Each simulated condition was replicated 5 times in WNWN and the paper
reports the mean and standard deviation of the replicate clustering
accuracies. The relatively small number of replicates accounts for the
jerkiness of the lines in the original Figures 4 and 5 (e.g.~the
Bray-Curtis distances calculated on the DEeqVS transformed data). A
better approach would have been to use 100 replicates as this would
reduce the dependency of the results on the random number generator's
seed. By increasing the number of replicates it was also possible to
compare the probability of falsely and correctly clustering samples from
the same and different treatment groups together (see points 9-11,
below). Because the accuracies are unlikely to be symmetric around a
mean at larger accuracy values the median and 95\% confidence intervals
or intraquartile range should have been reported. To test the effect of
increasing the number of replicates, I pulled apart the code in
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
into individual R and bash scripts that were executed using a Snakemake
workflow with the same Conda environment as above. This was necessary
since the number of simulated conditions because of this change
increased 20-fold. Such intense data processing was not practical within
a single R-flavored markdown document. Again, the observed results were
qualitiatively similar to those generated using the single R-flavored
markdown file (Figures S4 (\textbf{pam\_subsample15\_fig\_4.pdf}) and
S5(\textbf{pam\_subsample\_fig\_5.pdf}) ). The increased number of
replications resulted in smoother lines and allowed me to present
empirical 95\% confidence intervals. For all analyses in the remainder
of this paper, I used 100 randomized replicates per condition.

@@

\hypertarget{deseq-based-variance-stabilization-was-used-with-inappropriate-distance-calculation-methods}{%
\subsubsection{3. DeSeq-based variance stabilization was used with
inappropriate distance calculation
methods}\label{deseq-based-variance-stabilization-was-used-with-inappropriate-distance-calculation-methods}}

Close comparison of the original Figure 4 and my version
(\textbf{pam\_subsample15\_fig\_4.pdf}) revealed one important
difference between the two plots. In the original analysis, the
accuracies for the Weighted UniFrac distances at the largest effect size
(i.e., 3.5) were 1.0 for median sequencing depths of 1,000, 2,000, and
10,000. In my version of the analysis, the values for the same
sequencing depths were 0.88, 0.89, and 1.00, respectively. The 95\%
confidence interval for these accuracies spaned 0.51 and 1.00. The
Bray-Curtis distances were also different by both methods at smaller
effect sizes and had wide confidence intervals. Inspection of the DeSeq
normalized OTU counts revealed that the method resulted in negative
values. In fact, rendering the R-flavored markdown files in WNWN's
Protocol S1 generated a warning message when passing the DeSeq
normalized counts to the Bray-Curtis calculator, which said, ``results
may be meaningless because data have negative entries in method
`bray'\,''. Although the weighted UniFrac calculator function did not
generate a similar warning message, negative values would also result in
similarly meaningless distances. Both are due to the fact that the
calculators sum the counts of each OTU in both samples being compared.
In contrast, a Euclidean distance does not use a similar sum, but sums
the square of the difference between the OTU abundance in each sample.

To assess the prevalence of negative counts in the simulated data, I
quantified the fraction of negative values in the OTU matrix from each
simulation and counted the number of simulations where the transformed
OTU table had at least one negative value (Figure
\textbf{deseq\_negative\_value.pdf}). In general the fraction of
negative OTU counts increased with effect size, but decreased with
sequencing effort. The fraction of simulations with at least onee
negative value increased with effect size and sequencing effort. The
high frequency of negative OTU counts resulted in highly variable
Bray-Curtis and weighted UniFrac values. It is likely that because the
WNWN analysis only used 5 replicates that the large variation in
accuracies at high effect sizes was missed initially. For the rest of
this reanalysis study, I will only report results using the DeSeq-based
variance stabilization transformation with the Euclidean distance.

\hypertarget{a-single-subsampling-of-each-dataset-was-evaluated-rather-than-using-rarefaction}{%
\subsubsection{4. A single subsampling of each dataset was evaluated
rather than using
rarefaction}\label{a-single-subsampling-of-each-dataset-was-evaluated-rather-than-using-rarefaction}}

@@

\hypertarget{subsampling-removed-the-smallest-15-of-the-samples}{%
\subsubsection{6. Subsampling removed the smallest 15\% of the
samples}\label{subsampling-removed-the-smallest-15-of-the-samples}}

In WNWN, the authors quantified the tradeoff between sampling effort,
the number of samples removed below the threshold, and clustering
accuracy (original Figure 5, my Figure S5). Although the optimal
sampling effort varied by distance metric, transformation method, and
sampling effort, they removed samples whose number of sequences was less
than the 15th percentile (L404-419). They acknowledged that this
screening step, which was only used with subsampling, would decrease
clustering accuracy putting it at a relative disadvantage to the other
methods (page 5, column 1, last paragraph). Therefore, it was not
surprising that the peak clustering accuracy for their subsampled data
was at 85\% (Figure S2). Because the true result would not known \emph{a
priori} in microbiome studies, it would be impossible for researchers to
conduct a sensitivity analysis comparing the tradeoffs between
sequencing depth, sample number, and clustering accuracy to select a
sampling depth. {[}\textbf{Although the authors claim that ``Rarefying
counts requires an arbitrary selection of a library size minimum that
affects downstream inference'' (page 8, column 1, point 3), in acutal
microbiome studies the selection of a sampling depth is not as arbitrary
as the authors claim. Rather, to avoid p-hacking, researchers pick a set
of criteria where they will include or exclude samples prior to testing
their data.}{]} The differences in clustering accuracy between
subsampling and rarefaction and using PAM and K-means clustering
indicated that it was necessary to reassess the tradeoff between the
library size minimum quantile and clustering accuracy. When using
rarefaction, K-means clustering, and only considering conditions with
2,000 or more sequences, there was not a condition where setting a
higher threshold resulted in a better accuracy than using all of the
samples (\textbf{kmeans\_rarefaction\_fig\_5.pdf}). These results showed
that for modern sequencing depths, using the full datasets with
rarefaction and K-means clustering resulted in accuracies that were
typically better than those observed when removing the smallest 15\% of
the samples from each simulated dataset. When the original Figure 4 was
recast with these approaches, Rarefaction performed at least as well as
any of the other tranformations with each distance calculation, except
for with the Poisson distance (Figure
\textbf{kmeans\_rarefaction15\_fig\_4.pdf}). It is worth noting that at
larger effect sizes, K-means clustering did not perform as well for some
combinations of normalization methods and distance calculations (compare
\textbf{kmeans\_rarefaction15\_fig\_4.pdf} and
\textbf{pam\_subsample00\_fig\_4.pdf}); however, those combinations that
performed worse by K-means were not as good as rarefaction or
subsampling by either clustering method.

\hypertarget{the-distribution-of-library-sizes-was-poorly-chosen}{%
\subsubsection{7. The distribution of library sizes was poorly
chosen}\label{the-distribution-of-library-sizes-was-poorly-chosen}}

As described above, the sequencing depths used in the 26 GlobalPatterns
datasets were used as the distribution to create sequencing depths for
the 80 samples that were generated in each simulation. The
GlobalPatterns datasets had a mean of 1,085,256.85 sequences and a
median of 1,106,849.00 sequences per dataset. The datasets ranged in
sequencing depth between 58,688.00 and 2,357,181.00 sequences for a
40.16-fold difference. Rather than representing a typically observed
distribution of sequencing depths that would be skewed right, the
sampling distribution was normally distributed (Shapiro-Wilk test of
normality, P=0.57) (Figure \textbf{\emph{distribution\_shape.pdf}}).
From these simulations it is unclear how sensitive the various
transformations and distance calculations are to a skewed distribution.
A second limitation of this sampling distribution is that it only
contained 26 unique sampling depths such that each sampling depth would
have been re-used an average of 3.08 times in each simulation. Yet, it
is unlikely for a real sequence collection to have duplicate sequencing
depths.

>>>>

\hypertarget{waste-not-want-not-revisiting-the-analysis-that-called-the-practice-of-rarefying-microbiome-data-into-question}{%
\section{Waste not, want not: Revisiting the analysis that called the
practice of rarefying microbiome data into
question}\label{waste-not-want-not-revisiting-the-analysis-that-called-the-practice-of-rarefying-microbiome-data-into-question}}

@@

Since the development of highly parallelized sequencing technologies
that enable microbiome researchers to distribute sequences across
multiple samples in a single sequencing run, researchers have struggled
to produce a consistent number of sequences from each sample in a
dataset. It is common to observe more than 10-fold variation in the
number of sequences per sample {[}XXXXX{]}. Researchers desire
strategies to limit uneven sampling effort and need methods to control
for he uneveness in their analyses. Of course, uneven sampling is not
unique to microbiome research and is a challenge faced by all community
ecologists. Common approaches to controlling uneven sampling efforts
include use of proportional abundance (i.e., relative abundance),
normalization of counts, parameter estimation, and rarefaction.

In 2014 Paul McMurdie and Susan Holmes published ``Waste not, want not:
why rarefying microbiome data is inadmissible'' (WNWN) in PLOS
Computational Biology {[}XXXXX{]}. This paper has had a significant
impact on the approaches that microbiome researchers use to analyze 16S
rRNA gene sequence data. According to Google Scholar, this paper has
been cited more than 2,400 times as of April 2023. Anecdotely, I have
received correspondence from researchers over the past 10 years asking
how to address critiques from reviewers who criticize my correspondents'
analysis for using rarefaction (e.g., see
\href{https://twitter.com/inanna_nalytica/status/1264299305067786242}{this
Twitter thread}). I have also received similar comments from reviewers
regarding my own work. Most recently, I received the critique for a
preprint that I posted in 20XX that analyzes the practice of removing
rare taxa from microbiome analyses {[}XXXXX{]}. In the process of
responding to these reviewers' comments and preparing a manuscript
investigating rarefaction and other approaches to control for uneven
sequencing effort, I decided to reassess the WNWN study including their
definitions, simulations, and analyses.

@@

As I attempted to reproduce the results of WNWN, I noticed that the step
that rarefied the data only performed one subsampling of the data. This
caused me to re-inspect how McMurdie and Holmes defined ``rarefying'' in
the following quoted text from their paper:

@@

It was unfortunate that McMurdie and Holmes used the term ``rarefying''
here and throughout their manuscript. The authors were correct to state
that the distinction between ``rarefying'' and ``rarefaction'' is
confusing and leads to their conflation. Adding to the confusion is that
the papers cited in the first sentence of this quote (i.e., their
references {[}17{]}--{[}19{]}) either do not use the words ``rarefy'' or
``rarefying'' or use them interchangably with ``rarefaction''. In my
experience, subsequent researchers have conflated the results of this
study of the effects of rarefying data with rarefaction of data. As an
example, Willis (XXXXX) describes problems with rarefaction rather than
rarefying data when citing WNWN in her paper proposing alternatives to
rarefaction for use with alpha diversity data:

@@

In hindsight, as shown in the quoted text from WNWN, McMurdie and Holmes
did emphasize the distinction between rarefying and rarefaction.
However, because they seem to have coined a new meaning for rarefying,
they added to the confusion by using the generally used verb form of
rarefaction. Further confusion comes from the author's admonition in the
final sentence that some researchers have failed to repeat the
subsampling step. To most scientists, repeating the subsampling step is
rarefaction. My preference is to use \emph{subsampling} as the term
describing the process they refer to as rarefying. In other words,
subsampling is rarefaction, but with a single randomization. To minimize
confusion, I will use subsamping in place of rarefying.

I propose the following definition of rarefaction:

@@

This definition aligns well with how rarefaction was originally defined
for comparing richness (i.e., the number of taxa in a community) across
communities when communities are sampled to different depths
{[}XXXXXXXXX{]}. With this more general approach to rarefaction,
rarefaction can be performed using any alpha or beta diversity metric.
This strategy has been widely used by my research group and others and
is available in the mothur software package using commands such as
\texttt{summary.single}, \texttt{rarefaction.single},
\texttt{dist.shared}. The procedure outlined above could also be used
for hypothesis tests of differential abundance; however, thought would
need to be given to how to synthesize the results of these tests across
a large number of replications.

@@

McMurdie and Holmes analyzed the effect of rarefying and other
approaches on clustering accuracy using what they called ``Simulation
A'' in their Figure 2A and elsewhere in their paper. In Simulation A,
they investigated the ability to correctly assign simulated microbiome
samples to one of two clusters representing two simulated treatment
groups. Thankfully, the R code for Simulation A was provided by the
authors in the
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
R-flavored markdown file that was published as Protocol S1 in the
original paper. I will outline the simulation strategy below and will
reference line numbers from their R-flavored markdown file with ``L'' as
a prefix.

For the WNWN cluster analysis, sampling distributions for the two
treatment groups were generated using human fecal and ocean data
originally take from the GlobalPaterns dataset (L129) {[}XXXXX{]}. To
generate a fecal and ocean distribution, the authors included any
operational taxonomic unit (OTU) that appeared in more than one of the 4
fecal and 3 ocean samples (L60 and L137). The OTUs were sorted by how
many of the 7 samples the OTUs were observed in followed by their total
abundance across all 7 samples (L139). From this sorted list they
identified the identifiers of the first 2000 OTUs (L66). Returning to
the 7 samples they selected the data for the corresponding 2000 OTUs and
pooled the OTU abundances of the fecal and ocean samples separately to
create two sampling distributions (L144, L159-160, L197-198). Next, the
fecal and ocean distributions were mixed in 8 different fractions to
generate two community types that differend by varying effect sizes
(i.e., 1, 1.15, 1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220); an
effect size of 1 generated a null model with no difference between the
treatment groups. To simulate the variation in sequencing depth across
the 80 samples, they normalized he number of sequences from each of the
26 samples in the GlobalPatterns dataset so that the median number of
sequences (N\textsubscript{L}) for the GlobalPatterns had 1,000, 2,000,
5,000, or 10,000 sequences (L324-325). They then randomly sampled the 26
normalized sequencing depths to generate 80 sampling depths. From each
community type, they simulated 40 samples by sampling to the desired
number of sequences (L73, L230-233 and L326-327). Each simulation
condition was repeated 5 times (L85). This resulted in 160 simulations
(8 effect sizes x 4 median sampling depths x 5 replicates = 160
simulations). Finally, they removed rare and low prevalence OTUs in two
steps. First, they removed any OTUs whose total abundance was less than
3 across all 80 samples and that did not appear in at least 3 samples
(L368-386). Second, they removed any OTUs that did not have more than 1
sequence in more than 5\% of the 80 samples (i.e., 4 samples) and that
did not have a total abundance across the 80 samples greater than one
half of the number of samples in each community type (i.e., 20)
(L523-538, L551).

After generating the OTU counts for the 160 simulated communities, the
authors applied several normalization methods, distance calculations,
and clustering algorithms to the data. To normalize the OTU data the
original analysis either applied no normalization procedure, calculated
OTU relative abundances, subsampled the data, performed variance
stabilization using the DESeq R package, or performed Upper-Quartile
log-fold change normalization using the edgeR R package. In subsampling
the data, the authors either included all of the samples or removed
samples whose sequencing depth fell below the 5, 10, 15, 20, 25, and 40
percentile across all 80 samples. Subsampled data were then used as
input to calculate distances between samples using Bray-Curtis,
Euclidean, Unweighted UniFrac, and Weighted Unifrac distances as
implemented in the Phyloseq R package, Poisson distance as implemented
in the PoiClaClu R package, and top-mean squared difference as
implemented in the edgeR R package. Un-normalized data were used to
calculate Bray-Curtis, Euclidean, Poisson, and Weighted UniFrac
distances. Relative abundance data were used to calculate Bray-Curtis,
Unweigthed UniFrac, and Weighted UniFrac distances. DESeq variance
stabilization normalized data were used to calculate Bray-Curtis,
Euclidean, and Weighted Unifrac distances. The Upper-Quartile log-fold
change normalized data were only used to calculate top-mean squared
difference distances. The resulting distance matrices were used to
cluster the 80 samples into one of two clusters using partitioning
around the medoid (PAM), K-means clustering, and hierarchical
clustering. Although data for all three methods were presented in the
supplementary Protocol S1, only the PAM data are presented in the main
manuscript. The accuracy of the clustering assignments was quantified as
the fraction of the 80 samples that were assigned to the correct
cluster. Because some of the subsampling conditions removed samples
those were counted as mis-clustered samples yielding minimum accuracies
below 50\%.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Simulated conditions were only replicated 5 times each
\item
  The average sizes of the libraries were small by modern standards
\item
  DESeq-based variance stabilization was used with distance calculation
  methods that are sensitive to negative values
\item
  A single subsampling of each dataset was evaluated rather than using
  rarefaction, which likely resulted in noisier data
\item
  Results using PAM clustering were not directly compared to those of
  K-means and hierarchical clustering
\item
  Subsampling removed the smallest 15\% of the samples, which penalized
  accuracy values by 15 percentage points
\item
  The distribution of library sizes was not typical of those commonly
  seen in microbiome analyses
\item
  A filtering step was applied to remove rare taxa from the simulated
  datasets
\item
  No accounting of case when library sizes are confounded with treatment
  group
\item
  Clustering accuracy was used rather than direct comparisons of beta
  diversity
\item
  There was no consideration of effects of normalization methods on
  alpha diversity metrics
\end{enumerate}

@@

Each simulated condition was replicated 5 times in WNWN and the paper
reports the mean and standard deviation of the replicate clustering
accuracies. The relatively small number of replicates accounts for the
jerkiness of the lines in the original Figures 4 and 5 (e.g.~the
Bray-Curtis distances calculated on the DESeqVS normalized data). A
better approach would have been to use 100 replicates as this would
reduce the dependency of the results on the random number generator's
seed. By increasing the number of replicates it was also possible to
compare the probability of falsely and correctly clustering samples from
the same and different treatment groups together (see points 9-11,
below). Because the accuracies are unlikely to be symmetric around a
mean at larger accuracy values the median and 95\% confidence intervals
or intraquartile range should have been reported. To test the effect of
increasing the number of replicates, I pulled apart the code in
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
into individual R and bash scripts that were executed using a Snakemake
workflow with the same Conda environment as above. This was necessary
since the number of simulated conditions because of this change
increased 20-fold. Such intense data processing was not practical within
a single R-flavored markdown document. Again, the observed results were
qualitiatively similar to those generated using the single R-flavored
markdown file (Figures S4 (\textbf{pam\_subsample15\_fig\_4.pdf}) and
S5(\textbf{pam\_subsample\_fig\_5.pdf}) ). The increased number of
replications resulted in smoother lines and allowed me to present
empirical 95\% confidence intervals. For all analyses in the remainder
of this paper, I used 100 randomized replicates per condition.

@@

\hypertarget{deseq-based-variance-stabilization-was-used-with-distance-calculation-methods-that-are-sensitive-to-negative-values}{%
\subsubsection{3. DESeq-based variance stabilization was used with
distance calculation methods that are sensitive to negative
values}\label{deseq-based-variance-stabilization-was-used-with-distance-calculation-methods-that-are-sensitive-to-negative-values}}

Close comparison of the original Figure 4 and my version
(\textbf{pam\_subsample15\_fig\_4.pdf}) revealed one important
difference between the two plots. In the original analysis, the
accuracies for the Weighted UniFrac distances at the largest effect size
(i.e., 3.5) were 1.0 for median sequencing depths of 1,000, 2,000, and
10,000. In my version of the analysis, the values for the same
sequencing depths were 0.88, 0.89, and 1.00, respectively. The 95\%
confidence interval for these accuracies spaned 0.51 and 1.00. The
Bray-Curtis distances were also different by both methods at smaller
effect sizes and had wide confidence intervals. Inspection of the DESeq
normalized OTU counts revealed that the method resulted in negative
values. In fact, rendering the R-flavored markdown files in WNWN's
Protocol S1 generated a warning message when passing the DESeq
normalized counts to the Bray-Curtis calculator, which said, ``results
may be meaningless because data have negative entries in method
`bray'\,''. Although the weighted UniFrac calculator function did not
generate a similar warning message, negative values would also result in
similarly meaningless distances. Both are due to the fact that the
calculators sum the counts of each OTU in both samples being compared.
In contrast, a Euclidean distance does not use a similar sum, but sums
the square of the difference between the OTU abundance in each sample.

To assess the prevalence of negative counts in the simulated data, I
quantified the fraction of negative values in the OTU matrix from each
simulation and counted the number of simulations where the normalized
OTU table had at least one negative value (Figure
\textbf{deseq\_negative\_value.pdf}). In general the fraction of
negative OTU counts increased with effect size, but decreased with
sequencing effort. The fraction of simulations with at least onee
negative value increased with effect size and sequencing effort. The
high frequency of negative OTU counts resulted in highly variable
Bray-Curtis and weighted UniFrac values. It is likely that because the
WNWN analysis only used 5 replicates that the large variation in
accuracies at high effect sizes was missed initially. For the rest of
this reanalysis study, I will only report results using the DESeq-based
variance stabilization normalization with the Euclidean distance.

\hypertarget{a-single-subsampling-of-each-dataset-was-evaluated-rather-than-using-rarefaction-which-likely-resulted-in-noisier-data}{%
\subsubsection{4. A single subsampling of each dataset was evaluated
rather than using rarefaction, which likely resulted in noisier
data}\label{a-single-subsampling-of-each-dataset-was-evaluated-rather-than-using-rarefaction-which-likely-resulted-in-noisier-data}}

@@

\hypertarget{subsampling-removed-the-smallest-15-of-the-samples-which-penalized-accuracy-values-by-15-percentage-points}{%
\subsubsection{6. Subsampling removed the smallest 15\% of the samples,
which penalized accuracy values by 15 percentage
points}\label{subsampling-removed-the-smallest-15-of-the-samples-which-penalized-accuracy-values-by-15-percentage-points}}

In WNWN, the authors quantified the tradeoff between sampling effort,
the number of samples removed below the threshold, and clustering
accuracy (original Figure 5, my Figure S5). Although the optimal
sampling effort varied by distance metric, normalization method, and
sampling effort, they removed samples whose number of sequences was less
than the 15th percentile (L404-419). They acknowledged that this
screening step, which was only used with subsampling, would decrease
clustering accuracy putting it at a relative disadvantage to the other
methods (page 5, column 1, last paragraph). Therefore, it was not
surprising that the peak clustering accuracy for their subsampled data
was at 85\% (Figure S2). Because the true result would not known \emph{a
priori} in microbiome studies, it would be impossible for researchers to
conduct a sensitivity analysis comparing the tradeoffs between
sequencing depth, sample number, and clustering accuracy to select a
sampling depth. {[}\textbf{Although the authors claim that ``Rarefying
counts requires an arbitrary selection of a library size minimum that
affects downstream inference'' (page 8, column 1, point 3), in acutal
microbiome studies the selection of a sampling depth is not as arbitrary
as the authors claim. Rather, to avoid p-hacking, researchers pick a set
of criteria where they will include or exclude samples prior to testing
their data.}{]} The differences in clustering accuracy between
subsampling and rarefaction and using PAM and K-means clustering
indicated that it was necessary to reassess the tradeoff between the
library size minimum quantile and clustering accuracy. When using
rarefaction, K-means clustering, and only considering conditions with
2,000 or more sequences, there was not a condition where setting a
higher threshold resulted in a better accuracy than using all of the
samples (\textbf{kmeans\_rarefaction\_fig\_5.pdf}). These results showed
that for modern sequencing depths, using the full datasets with
rarefaction and K-means clustering resulted in accuracies that were
typically better than those observed when removing the smallest 15\% of
the samples from each simulated dataset. When the original Figure 4 was
recast with these approaches, Rarefaction performed at least as well as
any of the other tranformations with each distance calculation, except
for with the Poisson distance (Figure
\textbf{kmeans\_rarefaction15\_fig\_4.pdf}). It is worth noting that at
larger effect sizes, K-means clustering did not perform as well for some
combinations of normalization methods and distance calculations (compare
\textbf{kmeans\_rarefaction15\_fig\_4.pdf} and
\textbf{pam\_subsample00\_fig\_4.pdf}); however, those combinations that
performed worse by K-means were not as good as rarefaction or
subsampling by either clustering method.

\hypertarget{the-distribution-of-library-sizes-was-not-typical-of-those-commonly-seen-in-microbiome-analyses}{%
\subsubsection{7. The distribution of library sizes was not typical of
those commonly seen in microbiome
analyses}\label{the-distribution-of-library-sizes-was-not-typical-of-those-commonly-seen-in-microbiome-analyses}}

As described above, the sequencing depths used in the 26 GlobalPatterns
datasets were used as the distribution to create sequencing depths for
the 80 samples that were generated in each simulation. The
GlobalPatterns datasets had a mean of 1,085,256.85 sequences and a
median of 1,106,849.00 sequences per dataset. The datasets ranged in
sequencing depth between 58,688.00 and 2,357,181.00 sequences for a
40.16-fold difference. Rather than representing a typically observed
distribution of sequencing depths that would be skewed right, the
sampling distribution was normally distributed (Shapiro-Wilk test of
normality, P=0.57) (Figure \textbf{\emph{distribution\_shape.pdf}}).
From these simulations it is unclear how sensitive the various
normalizations and distance calculations are to a skewed distribution. A
second limitation of this sampling distribution is that it only
contained 26 unique sampling depths such that each sampling depth would
have been re-used an average of 3.08 times in each simulation. Yet, it
is unlikely for a real sequence collection to have duplicate sequencing
depths.

I created a new set of simulations that would focus on solely on the
effect of the shape of the distribution on the results to reassess the
WNWN results in the context of a more typical distribution of sample
sizes. I created a simple sequencing depth distribution where there were
80 depths logarithmically distributed between the minimum and maximum
sequencing depths of the GlobalPatterns dataset (Figure
\textbf{\emph{distribution\_shape.pdf}}). The median of this
distribution was 372,040.00 and the mean was 629,825.00. When I
regenerated the original Figures 4 and 5 using the log-distributed
sequencing effort distribution, the differences in normalization methods
were more apparent (Figure
\textbf{\emph{fig\_4\_kmeans\_rarefaction00\_log.pdf}}). For each of the
distance calculators, rarefying the data to the size of the smallest
dataset yielded accuracies that were at least as good as the other
methods across effect sizes and median sequencing depths. The difference
was most pronounced at smaller effect sizes and sequencing depths. When
comparing the performance of rarefaction across distance calculators for
different effect sizes, sequencing depths and size of smallest sample
(\textbf{\emph{fig\_5\_kmeans\_rarefaction\_log.pdf}}), the accuracies I
observed using the log-distributed sample sizes was at least as good as
those obtained using the GlobalPatterns-based distribution
\textbf{\emph{fig\_4\_kmeans\_rarefaction00\_a.pdf}}. The issue of
number of samples and the distribution of their sequencing depths in the
context of controlling for uneven sampling effort is explored in far
greater detail in an another analysis using sequencing depths observed
in biological samples {[}XXXXXX{]}.

\hypertarget{a-filtering-step-was-applied-to-remove-rare-taxa-from-the-simulated-datasets}{%
\subsubsection{8. A filtering step was applied to remove rare taxa from
the simulated
datasets}\label{a-filtering-step-was-applied-to-remove-rare-taxa-from-the-simulated-datasets}}
